SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2013-03-03 paper/paper.tex
paper updates.

\begin{abstract}
QuickCheck, developed by Klassen and Hughes, is a powerful library for automatic
test-case generation.  Because QuickCheck performs random testing, it can
discover very large counterexamples.  QuickCheck provides an interface for the
user to write \emph{shrink} functions to attempt to reduce the size of
counterexamples discovered.  Hand-written implementations of shrink can be
difficult to write: they can complex and consist of significant boilerplate
code.  Without a good understanding of QuickCheck and the property being tested,
they can be very inefficient, generating poor new candidates, and they may fail
to shrink a counterexample significantly.  In this paper, we introduce
\emph{SmartCheck}, an automatic efficient counterexample reduction algorithm.
SmartCheck reduces algebraic data using generic search heuristics to efficiently
find smaller counterexamples independent of the program or property being
tested.  In addition to shrinking, SmartCheck generalizes counterexamples by
returning a formula that abstracts over individual counterexamples and make it
easier for the programmer to uncover multiple counterexamples at once.
SmartCheck has been implemented for Haskell and is freely available.
\end{abstract}

@@

\todo{benchmark feat -- send to author}

@@

In industrial uses, shrinking is essential.  Hughes, in describing industrial
uses of QuickCheck, has noted that ``without it [shrinking], randomly generated
failing cases would often be so large as to be almost useless.''~\cite{qcjh}.
Hughes~\emph{et al.} give an extended example in which shrinking is essential in
debugging telecom software~\cite{telecom}.

Typically, effectively shrinking complex data requires a good understanding of
how shrinking in QuickCheck works and the semantics of the property and program
being evaluated.  Bad definitions can be so inefficient that they are unusable.
Shrinking presents a programmer's paradox: shrinking is used to generate a
smaller counterexample to better understand why a property failed, but to
implement an efficient shrink function often requires a deep enough
understanding of the property to efficiently generate smaller counterexamples to
it!

@@

To illustrate, consider the example in Figure~\ref{fig:initial}.\footnote{All
  examples and algorithms in this paper are presented in
  Haskell~\cite{haskell98}.  We strive to use idiomatic Haskell, and only rely
  on the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing four lists of
\ttp{Int16}s and a single \ttp{Word8}.  For our purposes, it is not important
what \ttp{T} models, but for concreteness, suppose it models the values in four
linked lists in an operating system, together with an 8-bit status register.

@@

We might first naively try to shrink counterexamples for a data type like
\ttp{T} by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- shrink w,  b <- shrink i0
                , c <- shrink i1, d <- shrink i2
                , e <- shrink i3 ]
\end{code}
%
\noindent
While the above definition appears reasonable, there are two problems with it.
First, the result of (\ttp{shrink t}) is null if any list contained in \ttp{t}
is null, then \ttp{t} will not be shrunk.  More troublesome is that with
QuickCheck's default \ttp{Arbitrary} instances, the length of potential
counterexamples returned by \ttp{shrink} can easily exceed $10^{10}$, which is
an intractable number of tests for QuickCheck to analyze.  The reason for the
blowup is that shrinking a list \ttp{[a]} produces a list \ttp{[[a]]}, the
length of which is significantly longer than the original list.  Then, we take
the cross-product of the generated lists.  For the example above, with some
counterexamples, the shrinking stage may appear to loop for hours, consuming
ever more memory, without returning a result.

@@

For a product type, SmallCheck must check values produced by taking the
cross-product of the type's fields at a given depth.  Unfortunately, SmallCheck
fails long before it reaches a depth required to discover a counterexample, due
to combinatorial explosion.  (A related library named Feat combines some aspects
of SmallCheck and QuickCheck~\cite{feat}; we discuss it more in the related work
(Section~\ref{sec:related}).

@@

Motivated by these limitations of QuickCheck and SmallCheck, we developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  SmartCheck takes a counterexample produced by some
oracle---in these experiments, we use QuickCheck---and then shrinks the
counterexample.

To motivate the benefit of SmartCheck, consider the graphs in
Figure~\ref{fig:graphs} and the Table~\ref{table:results}, giving the results
for using QuickCheck and SmartCheck on the program shown in
Figure~\ref{fig:initial} (we omit SmallCheck, since as noted, it is not even
feasible for the example).  The graphs show the number of results, out of 2000
test runs, with the specified size of the final counterexample (by counting
\ttp{N} constructors) or requiring the specified execution time.  We show two
execution time graphs, since the scale of time scales differ dramatically.  In
the table, we show the mean, median, median, and the results at the 95th
percentile values.  (The curves produced by these values are not necessarily
Gaussian, so we omit standard the deviation.)

The experiments show three approaches to shrinking with QuickCheck: no
shrinking, to provide a baseline, truncated shrinking, as described above, in
which each list is truncated to 10 elements, and tuple shrinking, also described
above.  Note that SmartCheck produces smaller counterexamples than the other
approaches, with a very small tail.  SmartCheck has a very small variance---note
the mean and median are nearly identical, meaning there are very few outliers.

More dramatic, however, are the differences
in execution time.  The execution times for SmartCheck include the initial
counterexample generation time required by QuickCheck; thus the time required by
SmartCheck to shrink a counterexample is on average just 0.2 seconds (0.10 -
0.08).  As can be seen in the graphs, SmartCheck's execution time is a linear
function (like QuickCheck, without shrinking).  Truncation and tuple shrinking
produce very long tails, with some counterexamples taking over a minute to
shrink.  SmartCheck is largely insensitive to the size of the original
counterexample provided.

In the last column are the results for SmartCheck, which includes the time
required by QuickCheck to generate the original counterexample.  The median time
required by SmartCheck to shrink is 0.02 seconds (SmartCheck's execution time
minus the execution time for QuickCheck without shrinking), and the median size
of a resulting value is six \ttp{N} constructors.

And we stress that SmartCheck achieves these results \emph{generically},
requiring no shrink definitions to be provided by the user.

@@

We focus on \emph{algebraic data types}~\cite{}.  Algebraic data types are
represented as ``sums of products''.  We sometimes refer to the elements of a
sum type as a \emph{variant} that is \emph{tagged} by a constructor.  For
example, the type
%
\begin{code}
Maybe a = Just a | Nothing
\end{code}
%
\noindent
contains two variants, tagged by the constructors \ttp{Just} and \ttp{Nothing},
respectively.

@@

class Arbitrary a => SubTypes a where
  size     :: a -> Size
  index    :: a -> Idx -> Maybe SubVal
  replace  :: a -> Idx -> SubVal -> a
  constr   :: a -> String
  constrs  :: a -> [String]
  baseType :: a -> Bool
  subVals  :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value,
\item \ttp{index} returns a \emph{sub-value} at an breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{baseType} returns whether the type of the value is an ``interesting
  type'', which generally means it is user-defined and has positive arity.
  We discuss the purpose of base types in Section~\ref{sec:base}.
\item \ttp{subVals} returns a tree of all non base-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

\section{Shrinking Data}\label{sec:shrinking}
In this section, we describe how to efficiently shrink data values.  Before we
begin, recall the basic idea behind the \ttp{shrink} method of the
\ttp{Arbitrary} class: generate a list of values, each of which is smaller than
the current counterexample.  Each of the new values generated typically bears no
relationship to the original counterexample other than being smaller.

@@

The approach taken by \ttp{shrink} is the most naive graph search algorithm: it
simply generates a random set of potential vertexes to check in the graph.
However, it does not exploit that the original counterexample \ttp{cex} is
already in a neighborhood of other likely counterexamples.  So rather than
choose random vertexes, it may be more efficient to test other values that are
closely related to \ttp{cex}.  By restricting the search to \ttp{cex}'s
neighborhood, a smaller counterexample might be missed, and only a local minimum
is returned.  However, with a large set of values, finding a global minimum by
random sampling vertexes in the graph is possible but too inefficient, as we saw
from the example in the introduction.

@@

The algorithm is guaranteed to terminate: intuitively, the measure for the
function is that either the index increases or the size of the counterexample
being evaluated decreases.  The algorithm's complexity is
$\mathcal{O}(n^2)$, where $n$ is the number of constructors in the
counterexample, assuming that generating new sub-values and testing them is done
in constant time.

\subsection{Reduction Algorithm Details}
Having describe the reduction algorithm, there are two important details about
its design we describe below.

\subsubsection{Variant Counterexample Hypothesis}
A motivation for the design of the reduction algorithm is something we cal the
\emph{variant counterexample hypothesis}.  Recall that the algorithm begins at
the \emph{first} sub-value of the counterexample rather than the 0th sub-value.
The 0th sub-value of \ttp{cex} is \ttp{cex} itself.  No invariant of the
algorithm would be violated by beginning with the 0th sub-value, and in
particular, the algorithm would still terminate.  However, we begin with a
strict sub-value because of the hypothesis.  The hypothesis is that in the
search space of possible values from a given type \ttp{T}, starting with a
particular counterexample \ttp{cex}, we are more likely to find other
counterexamples from the same variant as \ttp{cex}.  For example, consider a
property about unintended variable capture over a language's parse tree
represented by a sum type with constructors for module imports, function
definitions, and global-variable assignments, respectively.  A function
definition counterexample can only be reduced to smaller function definition
counterexamples, the only construct in which variable capture is possible.

@@

Consequently, we distinguish a class of types referred to as \emph{base types}.
If the reduction algorithm encounters a base type, it is ignored.  By default,
this includes numeric types (e.g., \ttp{Int}, \ttp{Word}, \ttp{Float}, etc.,
\ttp{Bool}, and \ttp{Char}.  We also treat recursive container types like lists
specially: lists of base types are considered to be base types themselves.  This
prevents the algorithm from analyzing, for example, strings.  A list of a
non-base-types is shrunk by the algorithm.

@@

SmartCheck's implementation contains flags to allow the user to customize its
behavior.  Three flags that are relevant to the reduction algorithm are the
following:
%
\begin{code}
scMaxReduce :: Int
scMaxSize   :: Int
scMaxDepth  :: Maybe Int
\end{code}
%
\noindent
The \ttp{scMaxReduce} flag controls the number of values generated by the
reduction algorithm for each sub-value analyzed.  \ttp{scMaxSize} controls the
maximum size of values generated to replace sub-values by the reduction
algorithm.  Thus, new sub-values must be strictly smaller than the minimum of
\ttp{scMaxSize} and the size of the sub-value being replaced.  Finally,
\ttp{scMaxDepth} determines the maximum depth in the counterexample the
reduction algorithm should analyze.  A value of \ttp{Nothing} means that the
counterexample should be exhaustively reduced, as the algorithm is presented in
Figure~\ref{fig:reduction}.  For example, the depth of a list is determined by
its length, and the depth of the binary tree defined in Section~\ref{sec:reduct}
is determined by the function \ttp{depth}:
%
\begin{code}
depth L         = 0
depth (B t0 t1) = 1 + max (depth t0) (depth t1)
\end{code}
%
\noindent
Of the flags, \ttp{scMaxDepth} is the most important for controlling
efficiency, particularly for large product types with significant ``fan out''.
The number of sub-values of a value product type can grow exponentially with
respect to the depth.  Furthermore, note that as the reduction algorithm
descends further, there is less chance to reduce the size of the value overall,
since smaller and smaller sub-values are replaced.

@@

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
Recall from the introduction the quantified formula
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
characterizing a set of counterexamples.  Another example using the calculator
language from Section~\ref{sec:subval} is to generalize a counterexample like
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
\noindent
by
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
since any dividend results in divide-by-zero.  Not only do the generalizations
assist the programmer's insight, but they reduce the sheer size of the
counterexample.  We call the kind of formula just shown \emph{universal
  sub-value generalization} and is implemented in SmartCheck.

@@

\todo{explain scMaxForall}

@@

New sub-values are generated by \ttp{newVals}, shown in
Figure~\ref{fig:reduction}.  Here, we bound the size of values to generate by
the flag \ttp{scMaxSize}, which is independent of the size of the particular
sub-value.

@@

The extrapolation algorithm is unsound insofar as it extrapolations from a set
of counterexamples to a universal claim, similar to QuickSpec~\cite{qs}.  By
tuning the parameters, the risk of an unsound generalization is reduced.

@@

@@

Sum types denote choice in a data type.  Sometimes, a property over a sum type
fails because there is a bug for some of the variants but not others.  For
example, recall the calculator language from Section~\ref{sec:subval}.  The
property \ttp{prop\_div} fails only for values that contain a variant tagged
with the \ttp{Div} constructor.  Recall again the generalized counterexample
from Section~\ref{sec:generalization}:
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
Because the divisor does not generalize, we know there is something special
about it that causes failure.  But we might wonder if there is something special
about variants tagged by the \ttp{Add} constructor, or might we finding failing
sub-values with the other variants.

We therefore introduce another kind of generalization we call \emph{existential
  sub-value generalization}.  In this generalization, if there a counterexample
containing every possible variant as a sub-value, then we abstract it.  For
example, the following all characterize counterexamples:
%
\begin{code}
Div (C 7) (Add (C (-5)) (C 5))
Div (Add (C 3) (C 2)) (C 0)
Div (C (-2)) (Div (C 0) (C 1))
\end{code}
%
\noindent
Because there are only three variants in the type, we have found counterexamples
built from each of them in the divisor.  We therefore can claim the following
formula holds:
%
\begin{code}
forall values x .
  forall constructors C .
    there exist arguments \(\stackrel{\rightarrow}{y}\) .
      such that Div x (C\(\stackrel{\rightarrow}{y}\))
\end{code}
%
\noindent
We therefore present an existential sub-value generalization algorithm that
performs constructor generalization.  Like with the other algorithms, this
algorithm also performs a breadth-first search over a counterexample.

@@

The function \ttp{constrFail} returns true if we replace the sub-value at index
\ttp{idx} in counterexample \ttp{cex} with every possible variant given the type
and construct a counterexample to the property.  There are three guards to the
recursive function \ttp{constrFail'}: the first guard holds if the list of
constructors tagging variants in which a counterexample is found is equal in
size to the list of all possible constructors for the type.  The second guard
tests whether the set of test values is null; if so (and if the first guard
fails), then we have exhausted test values before finding all possible failing
variants.  Third, for a specific sub-value \ttp{v}, we test whether it fails the
property.  If so, we add its constructor to the list of constructors.  Note in
the definition of \ttp{prop'}, we add an additional precondition that the
current constructor is not an element of constructors already seen.  Thus,
(\ttp{go v}) returns (\ttp{Just True}) if (\ttp{replace cex idx v}) passes this
precondition (and any other preconditions of \ttp{prop}), but fails the
property.

@@

\subsection{Automated Precondition Strengthening}\label{sec:precondition}
The extrapolation algorithm generalizes a counterexample, but in the
``neighborhood'' of he original counterexample.  In particular, all
generalizations are from the same variant as the original counterexample.  To
help the programmer in the generalization step, we would also like a way to test
the property again, ensuring we get counterexamples (if they exist) outside of
the neighborhood of the original one.

@@

Figure~\ref{fig:cex-gen} illustrates the situation in which we have some values
and a property in the case that some values fail the property of the form
(\ttp{pre ==> prop}).  Our goal is to find additional counterexamples in the
shaded region, in which \ttp{pre} is satisfied, \ttp{prop} is not satisfied, and
the counterexample is not already characterized by a universal sub-value
generalization formula.  Iterating, we increasingly cover the shaded region
with new formulas until no more uncovered counterexamples exist or the testing
tool fails to find new counterexamples.

@@

The data type to be tested must derive the \ttp{Read}, \ttp{Show},
\ttp{Typeable}, and \ttp{Generic} type classes.  Deriving \ttp{Typeable} and
\ttp{Generic} require using the GHC language extensions \ttp{DeriveDataTypeable}
and \ttp{DeriveGeneric}, respectively.  \ttp{Typeable} is used for dynamic
typing in defining the \ttp{replace} method of \ttp{SubTypes} since it is
unknown at compile-time whether the value to be replaced and replacing value
have the same types.

Then, the user simply declares, for a data type \ttp{D},
%
\begin{code}
instance Subtypes D
\end{code}
%

SmartCheck does not implement a counterexample discovery algorithm.  While any
method can be used to discover an initial counterexample, SmartCheck by default
uses QuickCheck to discover an initial counterexample.  If QuickCheck is used,
instances for \ttp{Arbitrary} must be provided.

>>>>

\begin{abstract}
QuickCheck, developed by Klassen and Hughes, is a powerful library for automatic
test-case generation.  Because QuickCheck performs random testing, it can
discover very large counterexamples.  QuickCheck provides an interface for the
user to write \emph{shrink} functions to attempt to reduce the size of
counterexamples discovered.  Hand-written implementations of shrink can be
difficult to write: they may be complex, inefficient, and consist of significant
boilerplate code.  We introduce \emph{SmartCheck}, an automatic efficient
counterexample reduction algorithm.  SmartCheck reduces algebraic data using
generic search heuristics to efficiently find smaller counterexamples
independent of the program or property being tested.  In addition to shrinking,
SmartCheck generalizes counterexamples by returning a \emph{formula} that
abstracts over individual counterexamples and make it easier for the programmer
to uncover multiple counterexamples at once.  SmartCheck has been implemented
for Haskell and is freely available.
\end{abstract}

@@

@@

In industrial uses, shrinking is essential.  Hughes, in describing commercial applications of QuickCheck, has noted that ``without it [shrinking], randomly generated
failing cases would often be so large as to be almost useless.''~\cite{qcjh}.
Hughes~\emph{et al.} give an extended example in which shrinking is essential in
debugging telecom software~\cite{telecom}.

Typically, effectively shrinking complex data requires a good understanding of
how shrinking in QuickCheck works and the semantics of the property and program
being evaluated.  Bad definitions can be so inefficient that they are unusable.

@@

To illustrate, consider the example in Figure~\ref{fig:initial}.\footnote{All
  examples and algorithms in this paper are presented in
  Haskell~\cite{haskell98}.  We strive to use idiomatic Haskell, and only rely
  on the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing four lists of wrapped
\ttp{Int16}s and a single wrapped \ttp{Word8}.  For our purposes, it is not important
what \ttp{T} models, but for concreteness, suppose it models the values in four
linked lists in an operating system, together with an 8-bit status register.

@@

We might first naively try to shrink counterexamples for a data type like
\ttp{T} by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- shrink w,  b <- shrink i0
                , c <- shrink i1, d <- shrink i2
                , e <- shrink i3 ]
\end{code}
%
\noindent
While the above definition appears reasonable, there are two problems with it.
First, the result of (\ttp{shrink t}) is null if any list contained in \ttp{t}
is null, so \ttp{t} will not be shrunk.  More troublesome is that with
QuickCheck's default \ttp{Arbitrary} instances, the length of potential
counterexamples returned by \ttp{shrink} can easily exceed $10^{10}$, which is
an intractable number of tests for QuickCheck to analyze.  The reason for the
blowup is that shrinking a list \ttp{[a]} produces a list \ttp{[[a]]}, the
length of which is significantly longer than the original list.  Then, we take
the cross-product of the generated lists.  For the example above, with some
counterexamples, the shrinking stage may appear to execute for hours, consuming
ever more memory, without returning a result.

@@

For a product type, SmallCheck must check values produced by taking the
cross-product of the type's fields at a given depth.  Unfortunately, SmallCheck
fails long before it reaches a depth required to discover a counterexample.  (A
related library named Feat combines some aspects of SmallCheck and
QuickCheck~\cite{feat}; we discuss it in Section~\ref{sec:related}).

@@

Motivated by these limitations of QuickCheck and SmallCheck, we developed
\emph{SmartCheck}.  SmartCheck takes a counterexample produced by some
oracle---in these experiments, we use QuickCheck---and generically minimizes and
generalizes the counterexample, as described in the preceding bullet points.

To motivate the benefits of SmartCheck, consider the graphs in
Figure~\ref{fig:graphs} and the Table~\ref{table:results}, giving the results
for using QuickCheck and SmartCheck on the program shown in
Figure~\ref{fig:initial} (we omit SmallCheck, since as noted, it is not even
feasible for the example).  We graph counterexample size and execution time.
Out of 2000 runs, we show the number of runs resulting in a shrunk
counterexample of a given size (by counting \ttp{N} constructors), and we show
the number of runs executing within a given time limit (in seconds).  Note that
we provide two execution-time graphs, since the scale of time scales differ
dramatically.  In the table, we show the mean, median, median, and the results
at the 95th percentile values.  (The curves produced by these values are not
necessarily Gaussian, so we omit standard the deviation.)

The experiments show three approaches to shrinking with QuickCheck: no
shrinking, to provide a baseline, truncated shrinking, as described above, in
which each list is truncated to 10 elements, and tuple shrinking, also described
above.  Note that SmartCheck produces smaller counterexamples than the other
approaches, with a very small tail.

More dramatic, however, are the differences in execution time.  The execution
times for SmartCheck include the initial counterexample generation time required
by QuickCheck; thus the time required by SmartCheck to shrink a counterexample
is on average just 0.2 seconds (0.10 - 0.08).  As can be seen in the graphs,
SmartCheck's execution time scales linearly (like QuickCheck, without
shrinking).  Although it is difficult to tell from the graphs, truncation and
tuple shrinking produce very long tails, with some counterexamples taking over a
minute to shrink.  SmartCheck is largely insensitive to the size of the original
counterexample provided.

And we stress that SmartCheck achieves these results \emph{generically},
requiring no shrink definitions to be provided by the user.  Indeed, there is
nothing particularly special about the example given.

@@

We focus on \emph{algebraic data types}.  Algebraic data types are
represented as ``sums of products''.  We sometimes refer to the elements of a
sum type as a \emph{variant} that is \emph{tagged} by a constructor.  For
example, the type
%
\begin{code}
Maybe a = Just a | Nothing
\end{code}
%
\noindent
contains two variants, tagged by the constructors \ttp{Just} and \ttp{Nothing},
respectively.

@@

class Arbitrary a => SubTypes a where
  size     :: a -> Size
  index    :: a -> Idx -> Maybe SubVal
  replace  :: a -> Idx -> SubVal -> a
  constr   :: a -> String
  constrs  :: a -> [String]
  baseType :: a -> Bool
  subVals  :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at an breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{baseType} returns whether the type of the value is an ``interesting
  type'', which generally means it is user-defined and has positive arity.
  We discuss the purpose of base types in Section~\ref{sec:base}.
\item \ttp{subVals} returns a tree of all non base-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

\section{Shrinking Data}\label{sec:shrinking}
In this section, we describe how to efficiently and generically shrink algebraic data
values.  Before we begin, recall the basic idea behind the \ttp{shrink} method
of the \ttp{Arbitrary} class: generate a list of values, each of which is
smaller than the current counterexample.  Each of the new values generated
typically bears no relationship to the original counterexample other than being
smaller.

@@

The approach taken by typical \ttp{shrink} implementations is the most naive
graph search algorithm: it simply generates a random set of potential vertexes
to check in the graph.  However, it does not exploit that the original
counterexample \ttp{cex} is already in a neighborhood of other likely
counterexamples.  So rather than choose random vertexes, it may be more
efficient to test other values that are closely related to \ttp{cex}.  By
restricting the search to \ttp{cex}'s neighborhood, a smaller counterexample
might be missed, and only a local minimum is returned.  However, with a large
set of values, finding a global minimum by random sampling vertexes in the graph
is possible but too inefficient, as we saw from the example in the introduction.

@@

The algorithm is guaranteed to terminate: informally, the measure for the
function is that either the index increases or the size of the counterexample
being evaluated decreases.  The algorithm's complexity is
$\mathcal{O}(n^2)$, where $n$ is the number of constructors in the
counterexample, assuming that generating new sub-values and testing them is done
in constant time.

\subsection{Reduction Algorithm Details}
Having described the reduction algorithm, there are two important details about
its design we describe below.

\subsubsection{Variant Counterexample Hypothesis}
A motivation for the design of the reduction algorithm is something we call the
\emph{variant counterexample hypothesis}.  Recall that the algorithm begins at
the \emph{first} sub-value of the counterexample rather than the 0th sub-value.
The 0th sub-value of \ttp{cex} is \ttp{cex} itself.  No invariant of the
algorithm would be violated by beginning with the 0th sub-value, and in
particular, the algorithm would still terminate.  However, we begin with a
strict sub-value because of the hypothesis.  The hypothesis is that in the
search space of possible values from a given type \ttp{T}, starting with a
particular counterexample \ttp{cex}, we are more likely to find other
counterexamples from the same variant as \ttp{cex}.  For example, consider a
property about unintended variable capture over a language's parse tree
represented by a sum type with constructors for module imports, function
definitions, and global-variable assignments, respectively.  A function
definition counterexample can only be reduced to smaller function definition
counterexamples, the only construct in which variable capture is possible.

@@

Consequently, we distinguish a class of types referred to as \emph{base types}.
If the reduction algorithm encounters a base type, it is ignored.  By default,
this includes numeric types (e.g., \ttp{Int}, \ttp{Word}, \ttp{Float}, etc.,
\ttp{Bool}, and \ttp{Char}.  We also treat recursive container types like lists
specially: lists of base types are considered to be base types themselves.  This
prevents the algorithm from analyzing, for example, strings.  A list of
non-base-types, however, is shrunk by the algorithm.

@@

SmartCheck's implementation contains flags to allow the user to customize its
behavior.  Three flags that are relevant to the reduction algorithm are the
following:
%
\begin{code}
scMaxReduce :: Int
scMaxSize   :: Int
scMaxDepth  :: Maybe Int
\end{code}
%
\noindent
The \ttp{scMaxReduce} flag controls the number of values generated by the
reduction algorithm for each sub-value analyzed.  \ttp{scMaxSize} controls the
maximum size of values generated to replace sub-values by the reduction
algorithm.  Thus, new sub-values must be strictly smaller than the minimum of
\ttp{scMaxSize} and the size of the sub-value being replaced.  Finally,
\ttp{scMaxDepth} determines the maximum depth in the counterexample the
reduction algorithm should analyze.  A value of \ttp{Nothing} means that the
counterexample should be exhaustively reduced, as the algorithm is presented in
Figure~\ref{fig:reduction}.  For example, the depth of a list is determined by
its length, and the depth of the binary tree defined in Section~\ref{sec:reduct}
is determined by the function \ttp{depth}:
%
\begin{code}
depth L         = 0
depth (B t0 t1) = 1 + max (depth t0) (depth t1)
\end{code}
%
\noindent
Of the flags, \ttp{scMaxDepth} is the most important for controlling
efficiency, particularly for large product types with significant ``fan out''.
The number of sub-values of a product type value can grow exponentially with
respect to the depth.  Furthermore, note that as the reduction algorithm
descends further, there is less chance to reduce the size of the value overall,
since smaller and smaller sub-values are replaced.

@@

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
Recall from the introduction the quantified formula
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
characterizing a set of counterexamples.  Another example using the calculator
language from Section~\ref{sec:subval} is to generalize a counterexample like
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
\noindent
by
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
since any dividend results in divide-by-zero for the given divisor.  Not only do
the generalizations assist the programmer's insight, but they reduce the sheer
size of the counterexample.  We call the kind of formula just shown
\emph{universal sub-value generalization} and it is implemented in SmartCheck.

@@

@@

New sub-values are generated by \ttp{newVals}, shown in
Figure~\ref{fig:reduction}.  Like in the reduction algorithm, we bound the size
of values to generate by the flag \ttp{scMaxSize}, which is independent of the
size of the particular sub-value.  The flag \ttp{scMaxForall} is the analogue of
the \ttp{scMaxReduce} flag, determining the number of values generated in trying
to generalize a value.

@@

@@

\paragraph{Soundness}
The extrapolation algorithm is unsound in two ways.  First, it extrapolates from a set
of counterexamples to a universal claim, similar to QuickSpec~\cite{qs}.  By
tuning the parameters, the risk of an unsound generalization is reduced.
Second, note that each index is generalized with respect to the concrete
sub-values that exist in the counterexample.  So in some cases, a formula may be
returned that is overly general.  For example, consider the counterexample in
which both fields of \ttp{Add} contain values causing the failure:
%
\begin{code}
Add (Div (C 1) (C (-2)) (C 2))
    (Div (C 0) (C (-1)) (C 1))
\end{code}
%
\noindent
The universal generalization algorithm might return the formula
\noindent
%
\begin{code}
forall values x0 x1 . Add x0 x1
\end{code}
%
\noindent
The algorithm could be modified to so that when attempting to generalize a
sub-value at an index, all previously-generalized indexes are re-generalized
simultaneously, by taking the cross-product of random values for the current
index and for the previously-generalized indexes.  While this reduces the
potential for over-generalizing a counterexample, it is exponential in the
number of indexes generalized.  For universally quantified formulas, the reader
should read it as shorthand for quantifying each variable independently and
taking the conjunction of formulas.  For example, instead of
%
\begin{code}
forall values x0 x1 . Add x0 x1
\end{code}
%
\noindent
one should read
%
%
\begin{code}
      forall values x0 .
        Add x0 (Div (C 0) (C (-1)) (C 1))
  and forall values x1 .
        Add (Div (C 1) (C (-2)) (C 2)) x1
\end{code}

@@

Sum types denote choice in a data type.  Sometimes, a property over a sum type
fails because there is a bug for some of the variants but not others.  For
example, recall again the calculator language from Section~\ref{sec:subval}.  The
no-division-by-zero property fails only for values that contain a variant tagged
with the \ttp{Div} constructor.  Recall again the generalized counterexample
from Section~\ref{sec:generalization}:
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
Because the divisor does not generalize, we know there is something special
about it that causes failure.  But we might wonder if there is something special
about variants tagged by the \ttp{Add} constructor, or might we finding failing
sub-values with the other variants.

We therefore introduce another kind of generalization we call \emph{existential
  sub-value generalization}.  In this generalization, if there a counterexample
containing every possible variant as a sub-value, then we abstract it.  For
example, the following are all counterexamples:
%
\begin{code}
Div (C 7) (Add (C (-5)) (C 5))
Div (Add (C 3) (C 2)) (C 0)
Div (C (-2)) (Div (C 0) (C 1))
\end{code}
%
\noindent
Because there are only three variants in the type, we have found counterexamples
built from each of them in the divisor.  We therefore can claim the following
formula holds:
%
\begin{code}
forall values x .
  forall constructors C .
    there exist arguments \(\stackrel{\rightarrow}{y}\) .
      such that Div x (C\(\stackrel{\rightarrow}{y}\))
\end{code}
%
\noindent
We therefore present an existential sub-value generalization algorithm that
performs constructor generalization.  Like with the other algorithms, this
algorithm also performs a breadth-first search over a counterexample.

@@

The function \ttp{constrFail} returns true if we replace the sub-value at index
\ttp{idx} in counterexample \ttp{cex} with every possible variant given the type
and construct a counterexample to the property.  There are four guards to the
recursive function \ttp{constrFail'}: the first guard holds if the list of
constructors tagging variants in which a counterexample is found is equal in
size to the list of all possible constructors for the type.  The second guard
tests whether the set of test values is null; if so (and if the first guard
fails), then we have exhausted test values before finding all possible failing
variants.  Third, for a specific sub-value \ttp{v}, we test whether it fails the
property.  If so, we add its constructor to the list of constructors.
Otherwise, we simply recurse.  Note in the definition of \ttp{prop'}, we add an
additional precondition that the current constructor is not an element of
constructors already seen.  Thus, (\ttp{go v}) returns (\ttp{Just True}) if
(\ttp{replace cex idx v}) passes this precondition (and any other preconditions
of \ttp{prop}), but fails the property.

@@

\subsection{Automated Precondition Strengthening}\label{sec:precondition}
The universal and existential generalization algorithms generalize a
counterexample, but in the ``neighborhood'' of the original counterexample.  In
particular, all generalizations are from the same variant as the original
counterexample.  To help the programmer in the generalization step, we would
also like a way to test the property again, ensuring we get counterexamples (if
they exist) outside of the neighborhood of the original one.

@@

Figure~\ref{fig:cex-gen} illustrates the situation in which we have some values
and a property in the case that some values fail the property of the form
(\ttp{pre ==> post}).  Our goal is to find additional counterexamples in the
shaded region, in which \ttp{pre} is satisfied, \ttp{post} is not satisfied, and
the counterexample is not already characterized by a universal sub-value
generalization formula.  Iterating, we increasingly cover the shaded region
with new formulas until no more uncovered counterexamples exist or the testing
tool fails to find new counterexamples.

@@

The data type to be tested must derive the \ttp{Read}, \ttp{Show},
\ttp{Typeable}, and \ttp{Generic} type classes.  Deriving \ttp{Typeable} and
\ttp{Generic} require using the GHC language extensions \ttp{DeriveDataTypeable}
and \ttp{DeriveGeneric}, respectively.  \ttp{Typeable} is used for dynamic
typing in defining the \ttp{replace} method of \ttp{SubTypes} since it is
unknown at compile-time whether the value to be replaced and replacing value
have the same types.  (We only require \ttp{Read} because standard QuickCheck
does not return a counterexample; we read the shown value.)

Then, the user simply declares, for a data type \ttp{D},
%
\begin{code}
instance Subtypes D
\end{code}
%
Default instances are provided for common Prelude types and some additional
ones, including all types for which QuickCheck provides instances.
